---
title: "Data Structures Pt1"
output: learnr::tutorial
runtime: shiny_prerendered
---
```{r setup, include=FALSE}

library(learnr)
```
## Welcome!

You can write and execute code within this tutorial, *however* I recommended following the instructions and completing the challenges in Rstudio in order help you gain proficiency in navigating Rstudio.



## Create a data frame

Before we start examining data types, let's start with getting some data! 

We'll being by creating a data frame by creating a lists of lists (of vectors actually) using the `data.frame()` and combine `c()`  functions.

- We will create an object named `cats` and assign it the values withing our data frame using the `data.frame()` function. 
- Within this function we will name our different variables `coat`, `weight`, and `likes_strings`.
- Each of these variables will be assigned a list of values using the combine `c()` function. 

See example:


```{r catsdf,exercise= TRUE, eval=FALSE}
cats<- data.frame(coat = c("calico", "black", "tabby"),
                  weight = c(2.1, 5.0,3.2),
                  likes_string = c(1,0,1))
cats
```

## Challenge 1

**Challenge 1: Create a new data frame ** 

- Save the data frame as an object named `office`
- Inside the `data.frame()` function, you will list your variables and their values just like we did for the `cats` data frame.
-  Each variable will be assined a list of values using the `c` function. **Example:** `coats=c("tabby","calico")` in this example the variable `coats` has the two values- tabby and calico. 
- variables: `people`, `age`, `loves_beets`, `loves_jim`

- For `people` use the values : "Jim", "Pam", "Michael", "Dwight"

- For `age` use the values : 28, 26, 42, 35

- For `loves_beets` use the values: 0,1,0,1

- For `loves_jim` use the values : 1, 1, 1, 0

```{r df, eval=FALSE, exercise = TRUE, exercise.reveal_solution =TRUE}

```

**Syntax clue :
`objectname<-data.frame(variable1 =c(value1,value2),
variable2 = c(another_value,and_anothervalue))`**

<div id= "df-hint">
**Hint** Here is the code for a similar data frame:

```{r}
office<- data.frame(people = c("Creed", "Meredith"), age = c(29,40), loves_beets = c(1,1), loves_jim=c(0,0))
```

</div>

## Pull columns
```{r prepare-cats}
cats<- data.frame(coat = c("calico", "black", "tabby"),
                  weight = c(2.1, 5.0,3.2),
                  likes_string = c(1,0,1))
```


To pull a column from a data frame use the syntax `dataframe_name$column_name`.

Run this code, and consider the output. 

```{r pullcoats, eval = FALSE, exercise =TRUE, exercise.setup="prepare-cats"}
cats$coat

```

**Exercise: Pulling Collumns**
PUll the other two columns (`weight`, and `likes_string`) from the `cats` dataframe.

```{r pullexercise01,exercise = TRUE, exercise.setup="prepare-cats", eval =FALSE} 

```

## Challenge 2
```{r prepare-office}
office<- data.frame(people=c("Jim","Pam","Michael","Dwight"), age = c(28,26,42,35),loves_beets = c(0,1,0,1), loves_jim = c(1,1,1,0))
```

### Challenge 2a

**Challenge 2: Pulling Collumns**
Pull the `people` column from the `office` data frame. 

```{r pullofficecol, eval= FALSE, exercise = TRUE, exercise.setup = "prepare-office"}

```
<div id = "pullofficecol-hint">

To pull the `loves_beets` column from the `office` data frame use the following code:

```{r pullcolex, eval=FALSE, exercise = TRUE, exercise.setup = "prepare-office"}
office$loves_beets
```
</div>

Try pulling some of the other columns. 

```{r pullofficecol2, eval= FALSE, exercise = TRUE, exercise.setup = "prepare-office"}

```


## Operations on Columns

We can perform different operations on the values in the data frame. 

Run this chunk of code and take note of the output.

```{r columnop, eval=FALSE, exercise = TRUE, exercise.setup = "prepare-cats"}
cats$weight+2
```

Now consider the result when we call on the column `cats$weight` again. 

```{r checkcatweight, eval= FALSE, exercise = TRUE, exercise.setup = "prepare-cats"}
cats$weight
```

Is the output what you expected? 

How does the following code's results differ? Why?

```{r reassigncatweight, eval=FALSE, exercise = TRUE, exercise.setup="prepare-cats"}
cats$weight<- cats$weight +2
cats$weight

```


## Challenge 3

**Column Operations**

Add 2 years to everyone's age in the `office` data frame you created in challenge 1.**

```{r addtoage, eval=FALSE, exercise=TRUE, exercise.setup="prepare-office"}

```

<div id= "addtoage-hint">
**Hint:** First pull the `age` collum from the `office` data frame using the `$` symbol then add two to the colum using `+ 2 `.

</div>

**Paste() Function**

Use the `paste()` function to create a series of sentences that say each person age.

```{r pasteoffice, eval=FALSE, exercise = TRUE, exercise.setup="prepare-office"}

```

<div id= "pasteoffice-hint">
You will use a similar syntax as the example below:
```{r pastehint, eval=FALSE, exercise = TRUE, exercise.setup="prepare-cats"}
paste("My cat is a ", cats$coat, "and weighs", cats$weight)
```

</div>

## Data Types

**There are 5 main data types:**

- `double`: or `numeric`, all real numbers with or without decimals

- `integer`: real numbers without decimal, can use the suffix `L` to specify data as an integer

- `complex`: imaginary va
lue, use the suffix `i` to specify data as `complex`
- `logical`: also known as `boolean`; is either `TRUE` or `FALSE`

- `character`: a string or set of characters, use single quotes `' '` or double quotes `" "` to specify data as a characters

No matter how complicated our analyses become, all data in R is interpreted as one of these basic data types. This strictness has some really important consequences.

To check the data type of a value we can use the `typeof()` function. See the examples below:
```{r typeofexample, exercise = TRUE, eval=FALSE}
typeof(22.45)
typeof("hello")
typeof(FALSE)
```

**Exercise:** Try guessing the data type of each of the following values, then check to see if you are correct using the `typeof()` function.

- 3.14

- 1L

- 1+1i

- TRUE

- 'banana'

- cats$weight

```{r typeofexercise, exercise = TRUE, eval = FALSE, exercise.setup = "prepare-cats"}

```

## Resources and Citations
This tutorial was created as supplemental material to the McNair Data Science Series. 

The slides and materials were adapted directly from the [Software Carpentries](), specifically the [Data Structures workshop](https://swcarpentry.github.io/r-novice-gapminder/). 

Challenges and Learnr tutorial were created by Ernestine Brannon.
